---
title: "Homework 1"
author: "Aidan Gillespie"
date: "2026-09-24"
output: html_document
---

# Chapter 2: #1 all parts, #2 all parts, #3 all parts, #5, #6, #7 all parts, #8 all parts, #9 all parts, #10 all parts


## Question 1
### For each of parts (a) through (d), indicate whether we would generally expect the performance of a flexible statistical learning method to be better or worse than an inflexible method. Justify your answer.

### (a) The sample size $n$ is extremely large, and the number of predictors $p$ is small.

We would expect the performance a flexible model to be better because we have a lot of data and we have fewer predictors which leaves less room for overfitting. We would also be able to capture more complex relationships within the data with a more flexible model.

---

### (b) The number of predictors $p$ is extremely large, and the number of observations $n$ is small.

We would expect the performance of a flexible model to perform worse because it is a lot more likely that it overfits to the data since the number of predictors is large and the number of observations is small. It will naturally try to curve to each of the points and will be able to do so because of low sample size.

---

### (c) The relationship between the predictors and response is highly non-linear.

We would expect the performance of a flexible model to peform better than an inflexible model. Since the response is highly unlinear, it makes it hard for a inflexible (commonly linear) model to capture the relationship whereas a flexible model might be able to represent the curves of non linear data better.

---

### (d) The variance of the error terms, i.e., $\sigma^2 = \operatorname{Var}(\epsilon)$, is extremely high.

We would expect the performance of a flexible model to perform worse than an inflexible model. This is due to the high variance which the model will try to fit more closely. With a linear model, we could capture the underlying relationship better before the adding \epsilon, but the flexible model would try to account for the \epsilon a lot more.


## Question 2
### Explain whether each scenario is a classification or regression problem, and indicate whether we are most interested in inference or prediction. Finally, provide n and p.

### (a) We collect a set of data on the top 500 firms in the US. For each firm we record profit, number of employees, industry and the CEO salary. We are interested in understanding which factors affect CEO salary.

This is a regression problem because we are trying to use predictors to determine a continuous variable (CEO's salary), and this is more of an inference problem because we aren't looking for the exact prediction, but rather how the predictors affect the salary.

n = 500; p = 3

---

### (b) We are considering launching a new product and wish to know whether it will be a success or a failure. We collect data on 20 similar products that were previously launched. For each product we have recorded whether it was a success or failure, price charged for the product, marketing budget, competition price, and ten other variables.

This is a classification problem, and we are looking to predict. We would like concrete probability of success not an inference on how those variables affect it.

n = 20; p = 14

---

### (c) We are interested in predicting the % change in the USD/Euro exchange rate in relation to the weekly changes in the world stock markets. Hence we collect weekly data for all of 2012. For each week we record the % change in the USD/Euro, the % change in the US market, the % change in the British market, and the % change in the German market.

This is a regression problem and we are looking to predict values. It explicitly states we are predicting % change and it's regression because it is a continous variable and not a category.

n = 52; p = 4

---

## 3. We now revisit the bias-variance decomposition.

### (a) Provide a sketch of typical (squared) bias, variance, training error, test error, and Bayes (or irreducible) error curves, on a single plot, as we go from less flexible statistical learning methods towards more flexible approaches. The x-axis should represent the amount of flexibility in the method, and the y-axis should represent the values for each curve. There should be five curves. Make sure to label each one.

```{r}
```

---


### (b) Explain why each of the five curves has the shape displayed in part (a).

# ANSWER LATER

---

## 5. What are the advantages and disadvantages of a very flexible (versus a less flexible) approach for regression or classification? Under what circumstances might a more flexible approach be preferred to a less flexible approach? When might a less flexible approach be preferred?

The advantages of a very flexible model over a less flexible one is that it allows you to capture more complex relationships and usually have higher accuracy assuming large n and low p. The disadvantages are that it can overfit to data and generally training time takes longer. You would want to use a more flexible approach when dealing with a large amount of data and few predictors, and you'd want to use a less flexible approach when there is a low amount of data and many predictors or the variance of \epsilon is high.

---

## 6. Describe the differences between a parametric and a non-parametric statistical learning approach. What are the advantages of a parametric approach to regression or classification (as opposed to a non-parametric approach)? What are its disadvantages?

Parametric models take a specific form and have a fixed number of parameters, but non parametric models don't have a specific form and are more flexible to your output variable. The advantages of parametric models is they are less prone to overfit compared to non-parametric models. The advantages of non parametrics is they are very flexible and can capture more complex relationships. The disadvantages of parametric models is they can underfit due to not being able to capture as complex relationships. The disadvantages of non parametric models is they can overfit due to how flexible they are.

---

## 7. The table below provides a training data set containing six observations, three predictors, and one qualitative response variable.

### (a) Compute the Euclidean distance between each observation and the test point, X1 = X2 = X3 = 0.

```{r}
df <- data.frame(
	X1 = c(0,2,0,0,-1,1),
	X2 = c(3,0,1,1,0,1),
	X3 = c(0,0,3,2,1,1),
	Y = c("Red", "Red", "Red", "Green", "Green", "Red")
)

euclid_dist <- function(a,b) {
	return(sqrt(sum((a-b)^2)))
}

z = c(0,0,0)

for (i in 1:nrow(df)) {
    obs <- as.numeric(df[i, c("X1", "X2", "X3")])
    dist <- euclid_dist(z, obs)
    cat(obs, "->", dist, "\n")
}
```

---

### (b) What is our prediction with K = 1? Why?

Our prediction would be Green because the closest value is -1 0 1 which has a Y of Green.

---

### (c) What is our prediction with K = 3? Why?

Our prediction would be Red because the closest values are

```
-1 0 1 -> Green
 1 1 1 -> Red
 2 0 0 -> Red
```

and the majority of values are red.

---

### (d) If the Bayes decision boundary in this problem is highly non-linear, then whould we expect the best value for K to be large or small? Why?

We would expect the best value for K to be small. This is because the larger the K value the smoother of a boundary we see because it's averaging K data points. When K is smaller it allows for much more flexibility because we're using each individual point to decide whether it is Green or Red.

---

## 8. This exercise relates to the College data set, which can be found in the file College.csv on the book website. It contains a number of variables for 777 different universities and colleges in the US. (I will not be writing out questions here just answers)

### (a) load in data

```{r}
college <- read.csv("College.csv")
```

### (b)
```{r}
rownames(college) <- college[,1]
View(college)

college <- college[,-1]
View(college)
```

### (c)
```{r}
# (i)
summary(college)

# I get an error when I try to do this because
# X and Private are strings

# (ii)
pairs(college[,2:11])

# (iii)
# says it can't take a string variable so converting to a factor
college$Private <- as.factor(college$Private)
plot(Outstate ~ Private, data = college)

# (iv)
Elite <- rep("No", nrow(college))
Elite[college$Top10perc > 50] <- "Yes"
Elite <- as.factor(Elite)

college <- data.frame(college, Elite)
summary(college$Elite)
plot(Outstate ~ Elite, data = college)

# (v)
par(mfrow = c(2, 2))

hist(college$Apps, breaks = 5, main = "Applications", xlab = "Apps")
hist(college$Accept, breaks = 10, main = "Acceptances", xlab = "Accept")
hist(college$Outstate, breaks = 20, main = "Out-of-State Tuition", xlab = "Outstate")
hist(college$Grad.Rate, breaks = 40, main = "Graduation Rate", xlab = "Grad.Rate")

# note I like 20 breaks the most

# (vi)
# I like pairs, but I'd rather see pearson correlation numbers


plot(college$Apps, college$Accept)
college$AcceptanceRate <- (college$Accept / college$Apps) * 100
plot(college$Apps, college$AcceptanceRate)

hist(college$Outstate)

plot(college$Outstate, college$Room.Board)

numeric_college <- college[, sapply(college, is.numeric)]
round(cor(numeric_college),2)
```

I have learned that college applications and college acceptances have a pretty linear relationship, and I also took a look at the apps vs acceptance rate. Which I thought would be pretty negatively correlated because I figured people trying to get into the best schools would have a lot of applications but not a lot of acceptances; however it doesn't have as strong as a negative correlation as I thought it would. I also learned that room and board and out of state have a string positive correlation which makes sense.

## 9. This exercise involves the Auto data set studied in the lab. Make sure that the missing values have been removed from the data

### (a)
```{r}
cars <- read.csv("Auto.csv")
summary(cars)
```

The quantitative variables are mpg, cylinders, displacement, horsepower, weight, acceleration, year

The qualitative variables are origin and name.

### (b)
```{r}
range(cars$mpg) 
range(cars$cylinders) 
range(cars$displacement) 
range(cars$horsepower) # this shows as strings for some reason???
range(cars$weight) 
range(cars$acceleration) 
range(cars$year)
```

### (c)
```{r}

mean(cars$mpg); sd(cars$mpg)  
mean(cars$cylinders); sd(cars$cylinders)  
mean(cars$displacement); sd(cars$displacement)  
mean(cars$horsepower); sd(cars$horsepower)
mean(cars$weight); sd(cars$weight)  
mean(cars$acceleration); sd(cars$acceleration)  
mean(cars$year); sd(cars$year)
```

### (d)
```{r}
cars_rm <- cars[-c(10:85)]

range(cars_rm$mpg); mean(cars_rm$mpg); sd(cars_rm$mpg)  
range(cars_rm$cylinders); mean(cars_rm$cylinders); sd(cars_rm$cylinders)  
range(cars_rm$displacement); mean(cars_rm$displacement); sd(cars_rm$displacement)  
range(cars_rm$horsepower); mean(cars_rm$horsepower); sd(cars_rm$horsepower)
range(cars_rm$weight); mean(cars_rm$weight); sd(cars_rm$weight)  
range(cars_rm$acceleration); mean(cars_rm$acceleration); sd(cars_rm$acceleration)  
range(cars_rm$year); mean(cars_rm$year); sd(cars_rm$year)
```

### (e)
```{r}
cars_numeric <- cars[, c(
  "mpg", "cylinders", "displacement", 
  "weight", "acceleration", "year"
)]

round(cor(cars_numeric), 2)
plot(cars$displacement, cars$weight)
plot(cars$weight, cars$mpg)
```

It seems that displacement and weight have a very positive correlation while weight and mpg have a very negative correlation.

### (f)

I would justify that weight, displacement, and cylinders are best predictors. Since they have a strong negative correlation we should be able to say the lower those are the higher our mpg will be. My correlation matrix shows this.

## 10.

### (a)
```{r}
library(ISLR2)
# ?Boston tired of this printing out everytime I compile
```

This dataset has 506 rows and 13 cols.
crim -> crime per capita

‘crim’-> per capita crime rate by town.
‘zn’-> proportion of residential land zoned for lots over 25,000
‘indus’-> proportion of non-retail business acres per town.
‘chas’-> Charles River dummy variable (= 1 if tract bounds river; 0
‘nox’-> nitrogen oxides concentration (parts per 10 million).
‘rm’-> average number of rooms per dwelling.
‘age’-> proportion of owner-occupied units built prior to 1940.
‘dis’-> weighted mean of distances to five Boston employment centres.
‘rad’-> index of accessibility to radial highways.
‘tax’-> full-value property-tax rate per $10,000.
‘ptratio’-> pupil-teacher ratio by town.
‘lstat’-> lower status of the population (percent).
‘medv’-> median value of owner-occupied homes in $1000s.

### (b).

```{r}
pairs(Boston)
```

Crime and age seem to be positively correlated which makes sense but its showing like people in their 40s to 80s comitting a lot more crime than I would expect. The lower of the median value of the home it seems more crime is more likely. This one is kinda obvious but the number of rooms per dwelling correlates with the median value of the home.

### (c).

The predictors associated with per captia crime seem to be age, distance to the employment centers and median home value.

Age -> Positively correlated the older you are the more likely the per captia crime is higher.
Distance to employment center -> closer you are to employment center more likely there is higher crime per capita there.
Median home value -> the lower the value of the home the higher the crime per capita.

### (d)

```{r}
range(Boston$crim)
range(Boston$tax)
range(Boston$ptratio)
```

The crime rate ranges from 0.00632 to 88.9762, which suggests that some census tracts have high crime rates. Tax rates range from 187 to 711 which is a large difference in tax rates between census tracts. Pupil-teacher ratios range from 12.6 to 22.0 which seems pretty tame compared to the than the crime rates.

### (e) & (f)

```{r}
sum(Boston$chas == 1)
median(Boston$ptratio)
```

35 census tracts bound the Charles river, and the median pupil-teacher ratio is 19.05.

### (g)

```{r}
min(Boston$medv)
Boston[Boston$medv == min(Boston$medv), ]
summary(Boston)
```

The tract that is lowest is 399 and 406. 

‘crim’-> much higher than avg
‘zn’-> close to median and equal to avg
‘indus’-> slightly higher than median and mean
‘chas’-> close to median and mean
‘nox’-> in the 3rd quartile
‘rm’-> lower than median and mean
‘age’-> equal to the max
‘dis’-> much lower than mean and median in bottom 25%
‘rad’-> equal to the max
‘tax’-> in the upper 3rd
‘ptratio’-> in the upper 3rd
‘lstat’-> very close to the max
‘medv’-> equal to the min

It seems the homes are very old because many units built before 1940. They are very far from highways, and are very crime ridden. Seems like an abandoned building?

### (h)

```{r}
sum(Boston$rm > 7)
sum(Boston$rm > 8)
Boston[Boston$rm > 8, ]
```

All of these with more than 8 rooms per dwelling seem pretty old with low tax rate.


